{\tt APFELgrid}: a fast interface to hadronic observables for fits of parton densities

### 2016-04-30 paper.tex
updating figures
updating section 3

% \biboptions{}

@@

This interpolation strategy is implemented in the {\tt
  APPLgrid}~\cite{Carli:2010rw} and {\tt
  FastNLO}~\cite{Wobisch:2011ij} projects. For the computation of the hard
cross sections, these packages rely on external codes to which they
are interfaced by means of a suite of
functions allowing for the filling of PDF- and
$\alpha_s$-independent look-up tables of cross-section weights. Codes
such as {\tt MCFM}~\cite{Campbell:2010ff} and {\tt NLOJet++}~\cite{Nagy:2003tz} have been interfaced directly to 
 {\tt APPLgrid}/{\tt FastNLO} and more recently dedicated interfaces to
automated general-purpose Monte Carlo (MC) event generators have been
developed. The {\tt aMCfast} and {\tt MCgrid} codes
have been specifically designed to extract the relevant information
from the {\tt MadGraph5\_aMC@NLO}~\cite{Alwall:2014hca} and {\tt
  SHERPA}~\cite{Gleisberg:2008ta} event generators respectively.

@@

The {\tt FastKernel} method, developed in the framework of
the NNPDF global analyses~\cite{Ball:2010de}, was developed to address
this problem. This method differs from the standard procedure \`{a} la {\tt APPLgrid} or {\tt fastNLO} in that it
maximises the amount of information that is precomputed prior to fitting such as to minimise the amount
of operations to be performed during the fit. Specifically, the {\tt FastKernel} method relies on the precomputation
of the partonic hard cross sections and DGLAP evolution kernels
and their combination into a single look-up table, here called a {\tt FastKernel} ({\tt FK}) table, in such a way that the prediction for a given hadronic
observable can be obtained by a simple matrix product between
the respective {\tt FK} table and the initial scale PDFs.

In this paper we present the {\tt APFELgrid} package, a public implementation of the {\tt
  FastKernel} method in which the hard partonic cross sections provided in an {\tt APPLgrid} look-up table are combined with the DGLAP
evolution kernels provided by the {\tt APFEL}
package~\cite{Bertone:2013vaa}.

The outline of this paper is the following. In
Sect.~\ref{sec:FastKernel} we present the tecnical details of the
implemantation of the {\tt FastKernel} method.

@@

\subsection{The FastKernel method}
A number of tools, $e.g.$ {\tt
  APFEL}~\cite{Bertone:2013vaa} or {\tt HOPPET}~\cite{Salam:2008qg},
are available which perform PDF evolution via an analogous
procedure. {\tt APFEL} in particular, is able to provide the so-called
\textit{evolution operator}. The evolution operator, is computed
assuming that the DGLAP evolution can be encoded in a single operator
whose action on the intial scale PDFs proceduces the final scale
PDFs. Taking $Q_0$ as an initial scale and $Q_\tau$ as a final scale,
on the $x$-space grid one can write:
\begin{equation}\label{eq:fastPDFfinal_recalled}
  f_i(x_{\alpha},Q^2_\tau) = \sum_{k}
  \sum_\beta A^\tau_{\alpha\beta, ik}\;
  f_k(x_\beta,Q^2_0)\,, 
\end{equation}
where the operator $A$ is the solution of the DGLAP evolution
equation with baundary condition:
\begin{equation}
A^\tau_{\alpha\beta, ik} =
\delta_{\alpha\beta}\delta_{ik}\qquad\mbox{for}\qquad Q_\tau = Q_0\,.
\end{equation}

@@

\section{Performance benchmarks}

The {\tt APFELgrid} plugin generates {\tt FK} tables of the form expressed in Eq.~(\ref{eq:FKTable}) from weight grids supplied by {\tt APPLgrid} files. In this section we shall examine the performance differences between the two expressions for cross-section prediction Eq.~(\ref{eq:applconv}) ({\tt APPLgrid}) and Eq.~(\ref{eq:FK}) ({\tt FK}). For a comprehensive benchmark, we shall consider a wide range of processes including LHC and Tevatron electroweak vector boson production measurements, inclusive jet data and $t\bar{t}$ total cross-sections. Predictions are provided for a wide range of kinematics, for a total of 52 source {\tt APPLgrid} files roughly corresponding to the full collider dataset of the forthcoming NNPDF3.1 PDF determination.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[p]
  \centering
  \includegraphics[scale=0.5]{plots/t0a}
\caption{\small Performance comparisons between {\tt APFELgrid} with
  AVX, SSE3 and double precision convolution and {\tt APPLgrid}
  convolution time per point and process.}
\label{fig:benchmark}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[p]
  \centering                                      
  \includegraphics[scale=0.5]{plots/t0b}
\caption{\small Performance comparisons between {\tt APFELgrid} with
  AVX, SSE3 and double precision convolution and {\tt APPLgrid}
  convolution time per point and process.}
\label{fig:benchmark}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Conclusion}

@@

%% References with bibTeX database:

\bibliographystyle{elsarticle-num}
\bibliography{paper.bib}

>>>>

% \biboptions{}
\biboptions{numbers,sort&compress}

@@

This interpolation strategy is implemented in the {\tt
  APPLgrid}~\cite{Carli:2010rw} and {\tt
  FastNLO}~\cite{Wobisch:2011ij} projects. For the computation of the hard
cross-sections, these packages rely on external codes to which they
are interfaced by means of a suite of
functions allowing for the filling of PDF- and
$\alpha_s$-independent look-up tables of cross-section weights. Codes
such as {\tt MCFM}~\cite{Campbell:2010ff} and {\tt NLOJet++}~\cite{Nagy:2003tz} have been interfaced directly to 
 {\tt APPLgrid}/{\tt FastNLO} and more recently dedicated interfaces to
automated general-purpose Monte Carlo (MC) event generators have been
developed. The {\tt aMCfast} and {\tt MCgrid} codes
have been specifically designed to extract the relevant information
from the {\tt MadGraph5\_aMC@NLO}~\cite{Alwall:2014hca} and {\tt
  SHERPA}~\cite{Gleisberg:2008ta} event generators respectively.

@@

The {\tt FastKernel} method, developed in the framework of
the NNPDF global analyses~\cite{Ball:2010de}, was developed to address
this problem. This method differs from the standard procedure \`{a} la {\tt APPLgrid} or {\tt FastNLO} in that it
maximises the amount of information that is precomputed prior to fitting such as to minimise the amount
of operations to be performed during the fit. Specifically, the {\tt FastKernel} method relies on the precomputation
of the partonic hard cross-sections and DGLAP evolution kernels
and their combination into a single look-up table, here called a {\tt FastKernel} ({\tt FK}) table, in such a way that the prediction for a given hadronic
observable can be obtained by a simple matrix product between
the respective {\tt FK} table and the initial scale PDFs.

In this paper we present the {\tt APFELgrid} package, a public
implementation of the {\tt FastKernel} method in which the hard
partonic cross sections provided in an {\tt APPLgrid} look-up table
are combined with the DGLAP evolution kernels provided by the {\tt
  APFEL}
package~\cite{Bertone:2013vaa,Carrazza:2014gfa,Bertone:2015cwa,Bertone:2015lqa,Carrazza:2015dea,Bertone:2015gba}.

The outline of this paper is the following. In
Sect.~\ref{sec:FastKernel} we present the technical details of the
implementation of the {\tt FastKernel} method. Then, in
Sect.~\ref{sec:benchmark} we illustrate timings and performance
results of the {\tt APFELgrid} library and grids. Finally, in
Sect.~\ref{sec:conclusion} we conclude and discuss possible future
developments.

@@

\subsection{The FastKernel method}
A number of tools, $e.g.$ {\tt
  APFEL}~\cite{Bertone:2013vaa,Carrazza:2014gfa,Bertone:2015cwa,Bertone:2015lqa,Carrazza:2015dea,Bertone:2015gba}
or {\tt HOPPET}~\cite{Salam:2008qg}, are available which perform PDF
evolution via an analogous procedure. {\tt APFEL} in particular, is
able to provide the so-called \textit{evolution operator}. The
evolution operator, is computed assuming that the DGLAP evolution can
be encoded in a single operator whose action on the intial scale PDFs
proceduces the final scale PDFs. Taking $Q_0$ as an initial scale and
$Q_\tau$ as a final scale, on the $x$-space grid one can write:
\begin{equation}\label{eq:fastPDFfinal_recalled}
  f_i(x_{\alpha},Q^2_\tau) = \sum_{k}
  \sum_\beta A^\tau_{\alpha\beta, ik}\;
  f_k(x_\beta,Q^2_0)\,, 
\end{equation}
where the operator $A$ is the solution of the DGLAP evolution
equation with baundary condition:
\begin{equation}
A^\tau_{\alpha\beta, ik} =
\delta_{\alpha\beta}\delta_{ik}\qquad\mbox{for}\qquad Q_\tau = Q_0\,.
\end{equation}

@@

\section{Performance benchmarks}
\label{sec:benchmark}

The {\tt APFELgrid} plugin generates {\tt FK} tables of the form
expressed in Eq.~(\ref{eq:FKTable}) from weight grids supplied by {\tt
  APPLgrid} files. In this section we shall examine the performance
differences between the two expressions for cross-section prediction
Eq.~(\ref{eq:applconv}) ({\tt APPLgrid}) and Eq.~(\ref{eq:FK}) ({\tt
  FK}).
%
For a comprehensive benchmark, we consider here a wide range of
processes including LHC and Tevatron electroweak vector boson
production measurements, $t\bar{t}$ total cross-sections,
double-differential Drell-Yan cross-sections and inclusive jet
data. Predictions are provided for a wide range of kinematics, for a
total of 52 source {\tt APPLgrid} files roughly corresponding to the
full collider dataset of the forthcoming NNPDF3.1 PDF determination.

In Fig.~\ref{fig:timings} we compare the convolution time per point
between {\tt APFELgrid} and {\tt APPLgrid} for all the 52
processes. We show timings for the {\tt APFELgrid} library in four
different configurations: AVX-OpenMP 2x (2 CPUs), AVX SSE3, and
standard double precision dot product convolution. When comparing the
results we observe a great performance improvement for convolution
time per point when using {\tt APFELgrid}. In particular, when
comparing {\tt APFELgrid} AVX to {\tt APPLgrid} timings we obtain a
minimum of x10 improvement for electroweak vector boson production and
a maximum of x2000 for inclusive jet data. These results confirm the
great usefulness of {\tt APFELgrid} in terms of performance for fits
of PDFs.

In Fig.~\ref{fig:performance} we provide information about the {\tt
  APFELgrid} table generation, grid size and memory usage. In terms of
generation time per point, we observe a timings from few milliseconds
to 3.5 minutes per point. Such difference is almost due to the density
of weights contained in the input {\tt APFELgrid} requiring a larger
number of combination operations. When considering the grid size, we
confirm, as expected, that in several processes the {\tt APPLgrid}
files are smaller than {\tt APFELgrid} ones. This feature is justified
by the lack of compression of {\tt APFELgrid} files. Finally, when
measuring the resident set size memory used by the convolution
programs, the amount of memory required by {\tt APFELgrid} is smaller
than {\tt APPLgrid} for all processes considered here.

In conclusion, the great performance benefits obtained with {\tt
  APFELgrid} endorse the importance of this tool not only in PDF fits
but also in any application requiring a large number of convolutions.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
  \centering
  \includegraphics[scale=0.6]{plots/t0a}
\caption{\small Performance comparisons between {\tt APFELgrid} with
  AVX-OpenMP 2x, AVX, SSE3, double precision convolution and {\tt
    APPLgrid} convolution time per point and process.}
\label{fig:timings}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
  \centering                                      
  \includegraphics[scale=0.6]{plots/t0b}
\caption{\small {\tt APFELgrid} generation time per point and
  comparison to {\tt APPLgrid} for grid size and resident set size
  (RSS).}
\label{fig:performance}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Conclusion}
\label{sec:conclusion}

In this work we presented {\tt APFELgrid}, an efficient convolution
interface to hadronic observables for fits of parton densities.

@@

%% References with bibTeX database:
\section*{References}
\bibliographystyle{elsarticle-num}
\bibliography{paper}
